---
title: "IDA HW1 - Vignesh Murugan"
output: pdf_document
date: "2024-08-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ---------------------------------
# 1st question
# ---------------------------------

```{r}


# 1a
x <- c(3, 12, 6, -5, 0, 8, 15, 1, -10, 7)

#1b
y <- seq(from=min(x), to=max(x), length.out=10)

# 1c
sum(x)
sum(y)
mean(x)
mean(y)     
sd(x)
sd(y)
var(x)
var(y)
mad(x)
mad(y)
quantile(x, probs = seq(0, 1, 0.25))
quantile(y, probs = seq(0, 1, 0.25))
quantile(x, probs = seq(0, 1, 0.2))
quantile(y, probs = seq(0, 1, 0.2))

# 1d
z <- sample(x, size=7, replace=TRUE )
z

# 1e
t.test(x,y)
```
## Differences in means are not significant

```{r}
# 1f
x <- x[order(x)]
x
t.test(x,y ,paired=TRUE)

# 1g
negative_logical_vector <- x < 0
negative_logical_vector

# 1h
non_negative_logical_vector <- x >= 0
x <- x[non_negative_logical_vector]
x

```
# ---------------------------------
# 2nd question
# ---------------------------------

```{r}

# 2a
col1 <- c(1,2,3,NA,5)
col2 <- c(4,5,6,89,101)
col3 <- c(45,NA,66,121,201)
col4 <- c(14,NA,13,NA,27)
X <- rbind (col1,col2,col3,col4)

is.na(X)

X[!complete.cases(X),]

# 2b
y <- c(3,12,99,99,7,99,21)
y[y == 99] <- NA
y
sum(is.na(y))

```


# ---------------------------------
# 3rd question
# ---------------------------------


```{r}
# 3a
college <- read.csv("college.csv")
college <- as.data.frame(college)

# 3b
rownames(college) <- college[,1]

if (interactive()) {
  View(college)
}

college <- college[,-1]

# 3c
#i
summary(college)

college$Private <- as.factor(college$Private)

summary(college)

#ii
pairs(college[,1:10])

#iii
plot(college$Private, college$Outstate, xlab= "Private", ylab= "Outstate",
     main="Boxplots of Outstate versus Private")

# iv
Elite <- rep("No", nrow(college))
# Create a character vector of size 777 with all elements being "No" and assigns it to variable called Elite

Elite[college$Top10perc >50] <- "Yes"
# If the values in "Top10Perc" variable are greater than 50 then those indexes in Elite are turned to yes

Elite <- as.factor(Elite)
# we convert the character vector to factor vector

college <- data.frame(college ,Elite)
# creates a new variable in college called Elite

# v
summary(college)

# vi
plot(college$Elite, college$Outstate, xlab= "Elite", ylab= "Outstate",
     main="Boxplots of Outstate versus Elite")

# vii
par(mfrow=c(2,2))

hist(college$Accept, breaks=10, 
     main="Histogram of Accept with 10 Bins", xlab="Accept")

hist(college$Accept, breaks=20, 
     main="Histogram of Accept with 20 Bins", xlab="Accept")

hist(college$Accept, breaks=30, 
     main="Histogram of Accept with 30 Bins", xlab="Accept")

hist(college$Accept, breaks=40, 
     main="Histogram of Accept with 40 Bins", xlab="Accept")

```

# ---------------------------------
# 4th question
# ---------------------------------

```{r}
library(plyr)

# 4a
baseball <- baseball

# 4b
baseball$sf[baseball$year < 1954] <- 0

baseball$hbp[is.na(baseball$hbp)] <- 0

baseball <- baseball[baseball$ab >= 50,]

# 4c
baseball$obp <- with(baseball, (h + bb + hbp) / (ab + bb + hbp + sf))

# 4d
top_players <- baseball[order(-baseball$obp), 
                        c("year", "id", "obp")]
head(top_players, 5)
```

# ---------------------------------
# 5th question
# ---------------------------------

```{r}
# 5a
quakes <- quakes

# 5b
par(mfrow=c(1,1))
plot(quakes$mag, quakes$depth)

# 5c
quakeAvgDepth <- aggregate(depth ~ mag, data = quakes, mean)

# 5d
names(quakeAvgDepth)[2] <- "AvgDepth"
names(quakeAvgDepth)

# 5e
plot(quakeAvgDepth$mag,quakeAvgDepth$AvgDepth, 
     xlab = "magnitude of earthquakes", ylab = "average Depth")

```

## 5f

## From the plots we can say that as magnitude increases the depth decreases I believe there is a reason for this,so I spent sometime surfing the web.It seems that higher magnitude quakes on higher depths are less harmful towards earths surface.So You could say that the data only contains earthquakes which cause substantial damage to property on surface.


